---
title: "assignment_3"
format: html
editor: visual
---



## student A will work here -

## student B will work here -

```{r}
#| label: setup
#| include: false

## student C will work here -

#How have participation and medal-winning patterns changed across Olympic Games from 1896 to 2016, with a focus on gender representation and top-performing nations

```{r} Load required package}
#| label: get-data
#| message: false
#| include: false
library(tidyverse)
library(broom)
```

```{r}
#| label: load-data
#| include: false
data <- read_csv("data/athlete_events.csv")
```

## Methodology

How have participation and medal-winning patterns changed across Olympic Games from 1896 to 2016, with a focus on gender representation and top-performing nations

### Data source:
The dataset is sourced from Kaggle and acknowldeged as The Olympic data on www.sports-reference.com, which contains information on athletes, their participation in Olympic Games, and the medals they won from 1896 to 2016. The dataset includes variables:

- ID: Unique number for each athlete

- Name: Athlete's name

- Sex: M or F

- Age: Integer

- Height: In centimeters

- Weight: In kilograms

- Team: Team name

- NOC: National Olympic Committee 3-letter code

- Games: Year and season

- Year: Integer

- Season: Summer or Winter

- City: Host city

- Sport: Sport

- Event: Event

- Medal: Gold, Silver, Bronze, or NA

### Implementation:
To analyse how participation and medal-winning patterns have changed across the Olympic Games from 1896 to 2016, we examine four key aspects: overall participation, female representation, female medal success, and medal concentration among nations. Statistical models are used to assess whether these measures exhibit significant trends over time.

1. Overall participation patterns over time: 
To determine whether Olympic participation has increased over time, a linear regression model is fitted:
$Participation_{t}$ ~ $\beta_0$ + $\beta_1$ * Year + $\epsilon_t$
where $Participation_{t}$ is the number of athletes participating in each Year. A significant positive coefficient for Year ($\beta_1$) would indicate an increase in participation over time.

2. Female Representation over time:
To assess whether female representation has changed over time, the proportion of female athletes is analysed using a binomial logistic regression model:
$\log\left(\frac{p_{t}}{1 - p_{t}} \right)$ ~ $\beta_0$ + $\beta_1$ * Year + $\epsilon_t$
where $p_{t}$ is the proportion of female athletes in each Year. The model is fitted using the yearly counts of female and male athletes. A significantly positive estimate of $\beta_1$ would suggest an increasing trend in female representation over time

3. Female Medal Success over time:
To investigate whether female athletes have become more successful over time, the proportion of medals won by women is analysed using a binomial logistic regression model:
$\log\left(\frac{p_{t}}{1 - p_{t}} \right)$ ~ $\beta_0$ + $\beta_1$ * Year + $\epsilon_t$
where $p_{t}$ is the proportion of medals won by female athletes in each Year. The model is fitted using the yearly counts of medals won by female and male athletes. A significantly positive estimate of $\beta_1$ would indicate an increasing trend in female medal success over time.

4. Medal Concentration Among Nations:
To evaluate whether medal concentration among nations has changed over time, the Herfindahl-Hirschman Index (HHI) is calculated for each Year based on the distribution of medals among countries. A linear regression model is then fitted:
$HHI_{t}$ ~ $\beta_0$ + $\beta_1$ * Year + $\epsilon_t$
where $HHI_{t}$ is the Herfindahl-Hirschman Index for each Year. A significant negative coefficient for Year ($\beta_1$) would suggest a decrease in medal concentration among nations over time, indicating a more even distribution of medals across countries.

## Results

1. Overall participation patterns over time: 
```{r}
#| include: false
participation <- data %>%
  distinct(ID, Year, Season) %>%
  group_by(Year) %>%
  summarise(
    Athletes = n(),
    .groups = "drop"
  )

participation_model <- lm(Athletes ~ Year,
                          data = participation)
```

```{r}
#| echo: false
#| label: tbl-participation
summary(participation_model) %>% glance()
```

```{r}
#| label: fig-participation-residuals
#| fig-cap: "Residual plots for participation model"
#| fig-subcap:
#| - "Residuals vs Fitted"
#| - "Normal Q-Q"
#| echo: false
plot(participation_model, which = 1)
plot(participation_model, which = 2)
```

According to @tbl-participation, the coefficient for Year is positive and statistically significant (p<0.001), indicating that Olympic participation has increased over time. However, the relatively low R-squared value suggests that Year alone explains only a limited proportion of the variation in participation. Additionally, @fig-participation-residuals-1 and @fig-participation-residuals-2 show clear patterns rather than random variation, indicating that the assumptions of linear regression may not be fully satisfied. Therefore, while there is evidence of a significant upward trend in participation, a more complex model may better capture the relationship over time.

2. Female Representation over time:
```{r}
#| include: false
gender_participation <- data %>%
  distinct(ID, Year, Sex) %>%
  group_by(Year, Sex) %>%
  summarise(
    Athletes = n(),
    .groups = "drop"
  ) %>%
  tidyr::pivot_wider(
    names_from = Sex,
    values_from = Athletes,
    values_fill = 0
  ) %>%
  rename(
    Female = F,
    Male = M
  )

female_model <- glm(
  cbind(Female, Male) ~ Year,
  family = binomial,
  data = gender_participation
)
```


library(tidyverse)
library(knitr)
library(kableExtra)


athletes <- read_csv("data/athlete_events.csv")
```

## Methodology

This data analysis makes use of the "120 Years of Olympic History" dataset found on Kaggle and has a total of 271,116 observations in 15 different variables that cover the period from the Athens Games in 1896 to the Rio Games in 2016 [@olympics_kaggle]. Data is related to individual athletes such as their names, sex, age, country, sport, event, and the medals won. Only the Summer Olympic Games were considered in order to ensure consistency in the types of events being compared since there are differences between the Summer and Winter Olympic Games in terms of sports, countries that participate, and numbers of athletes. The Age, Height and Weight variables had missing values which were not necessary for the analysis. Medal values of NA indicate non-medallists and were excluded from medal counts. @fig-gender shows participation trends by gender over time and @tbl-medals summarises medal counts by sport for the top 5 countries.

```{r}
#| label: fig-gender
#| fig-cap: "Athlete Participation by Gender in the Summer Olympics (1896–2016)"
#| echo: false
#| warning: false

athletes |>
  filter(Season == "Summer") |>
  distinct(Year, Name, Sex) |>
  count(Year, Sex) |>
  ggplot(aes(x = Year, y = n, colour = Sex)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  labs(x = "Year", y = "Number of Athletes", colour = "Sex") +
  theme_minimal()

```

```{r}
#| label: tbl-medals
#| tbl-cap: "Top 5 Countries by Total Medals (Summer Olympics)"
#| echo: false
#| warning: false

athletes |>
  filter(Season == "Summer", !is.na(Medal)) |>
  count(Team, name = "Total Medals") |>
  arrange(desc(`Total Medals`)) |>
  head(5) |>
  rename(Country = Team) |>
  kable(booktabs = TRUE)
```

## Results

@fig-topcountries shows the ten countries with the highest total number of medals won through all Summer Olympic Games. Leading is the United States, then comes the Soviet Union and Germany. Australia can be found among the ten most successful countries. There is a noticeable domination of rich countries throughout history, which proves the correlation between national resources and Olympic success.

```{r}
#| label: fig-topcountries
#| fig-cap: "Top 10 Countries by Total Medal Count in the Summer Olympics (1896–2016)"
#| echo: false
#| warning: false

athletes |>
  filter(Season == "Summer", !is.na(Medal)) |>
  count(Team, name = "total_medals") |>
  arrange(desc(total_medals)) |>
  head(10) |>
  mutate(Team = fct_reorder(Team, total_medals)) |>
  ggplot(aes(x = total_medals, y = Team)) +
  geom_col(fill = "steelblue") +
  labs(x = "Total Medals", y = "Country") +
  theme_minimal()
```

## student C will work here -

```{r}
#| echo: false
#| label: tbl-female-representation
summary(female_model) 
```

```{r}
#| label: fig-female-representation-residuals
#| fig-cap: "Residual plots for female representation model"
#| fig-subcap:
#| - "Residuals vs Fitted"
#| - "Normal Q-Q"
#| echo: false
plot(female_model, which = 1)
plot(female_model, which = 2)
```

The binomial logistic regression indicates that female representation in the Olympic Games has increased significantly over time. Based on @tbl-female-representation, the coefficient for Year is positive and highly significant (p-value < 0.001) suggesting that the proportion of female athletes has steadily increased throughout the study period . The large reduction in deviance from the null model (20058.38) to the fitted model (503.82) indicates that Year explains a substantial amount of the variation in female participation rates. The residual plots look reasonablyrandom, indicating a good fit for the model. Therefore, there is strong evidence that female representation in the Olympics has increased over time.

3.
```{r}
#| include: false
medals <- data %>%
  filter(!is.na(Medal))
gender_medals <- medals %>%
  group_by(Year, Sex) %>%
  summarise(
    Medals = n(),
    .groups = "drop"
  ) %>%
  tidyr::pivot_wider(
    names_from = Sex,
    values_from = Medals,
    values_fill = 0
  ) %>%
  rename(
    FemaleMedals = F,
    MaleMedals = M
  )

female_medal_model <- glm(
  cbind(FemaleMedals, MaleMedals) ~ Year,
  family = binomial,
  data = gender_medals
)
```

```{r}
#| echo: false
#| label: tbl-female-medal-success
summary(female_medal_model) 
```

```{r}
#| label: fig-female-medal-residuals
#| fig-cap: "Residual plots for female medal success model"
#| fig-subcap:
#|  - "Residuals vs Fitted"
#|  - "Normal Q-Q"
#| echo: false
plot(female_medal_model, which = 1)
plot(female_medal_model, which = 2)
```

The logistic regression shows a significant increase in female medal success over time (p-value < 0.001). @tbl-female-medal-success shows the positive coefficient indicating that the proportion of medals won by female athletes has steadily increased throughout Olympic history. Furthermore, the substantial reduction in deviance from the null model (5548.17) to the fitted model (187.68) indicates that Year explains a considerable amount of the variation in female medal success rates. These results provide strong evidence that women's contribution to overall Olympic medal success has increased over time.

4. Medal Concentration Among Nations:
```{r}
#| include: false
country_medals <- medals %>%
  group_by(Year, NOC) %>%
  summarise(
    Medals = n(),
    .groups = "drop"
  )

hhi_data <- country_medals %>%
  group_by(Year) %>%
  mutate(
    Share = Medals / sum(Medals)
  ) %>%
  summarise(
    HHI = sum(Share^2),
    .groups = "drop"
  )

hhi_model <- lm(HHI ~ Year,
                data = hhi_data)
```

```{r}
#| echo: false
#| label: tbl-hhi
summary(hhi_model) %>% glance()
```

```{r}
#| label: fig-hhi-residuals
#| fig-cap: "Residual plots for HHI model"
#| fig-subcap: 
#|   - "Residuals vs Fitted"
#|   - "Normal Q-Q"
#| echo: false
plot(hhi_model, which = 1)
plot(hhi_model, which = 2)
```

@tbl-hhi indicates a significant decline in medal concentration over time. The coefficient for Year is negative and statistically significant with $\beta_1 = -0.00140$ and $p = 0.0017$, suggesting that Olympic medals have become less concentrated among a small number of countries and more evenly distributed across participating nations. The model explains approximately 26.2% of the variation in the HHI ($R^2 =0.262$), indicating that Year is a meaningful predictor of medal concentration, although other factors may also influence the distribution of medals. The residual plots satisfy the assumptions of linear regression, showing no clear patterns and suggesting a good fit for the model. Overall, these results provide strong evidence that medal concentration among nations has decreased over time, indicating a more competitive and inclusive Olympic Games.

