---
title: "120 Years of Olympic History: Participation and Medal Trends"
author: ["Vibhor Dabas", "Shravani Jadhav", "Minh Ngo"]
date: today
format:
  beamer: default
---

```{r}
#| label: load-packages
#| include: false
library(knitr)
library(ggplot2)
library(dplyr)
options(width = 70)
```

```{r}
#| label: get-data
#| message: false
#| include: false

library(tidyverse)
library(broom)
library(knitr)
library(kableExtra)
data <- read_csv("data/athlete_events.csv")
adj_data <- data %>% filter(Season == "Summer") 
adj_data_c <- data %>% filter(Season == "Summer") %>% mutate(Year = Year - min(Year))
```

## Introduction

### Olympic Games:

-   The most popular international sports competition every four years, with over 200 nations participate, with over 10,000 athletes, and over 300 events

-   Have expanded enormously in scale, diversity and global importance

Studying historical trends in Olympic participation and performance can shed light on wider trends in gender equality, national development and sporting investment. Analysed a data-based view on the evolution of the Olympic movement.

## Dataset + Research Questions

Data of athletes, events, medals, and attributes (Summer and Winter Olympics), collected from Kaggle, from 1896 to 2016 (120 years of Olympic history).

-   How has overall athlete participation in the Summer Olympic Games changed over time?

-   How has female representation among Olympic athletes changed over time?

-   Has the proportion of Olympic medals won by female athletes increased over time?

-   Has the distribution of Olympic medals among nations become more or less concentrated over time?

## Methodology

**Visualisation** of trends in participation and medals over time

**Implentation** of a simple linear regression model, logistic regression model and HHI to test specific hypotheses about the relationship

1. Overall participation patterns over time:
- $Participation_{t}$ ~ $\beta_0$ + $\beta_1$ * Year + $\epsilon_t$

2. Female Medal Success over time:
- $\log\left(\frac{p_{t}}{1 - p_{t}} \right)$ ~ $\beta_0$ + $\beta_1$ * Year + $\epsilon_t$

3. Medal Concentration Among Nations:
- $HHI_{t}$ ~ $\beta_0$ + $\beta_1$ * Year + $\epsilon_t$

## Results
### Visualisation

::: columns
:::: column
```{r}
#| label: fig-gender
#| fig-cap: "Athlete Participation by Gender in the Summer Olympics (1896–2016)"
#| echo: false
#| warning: false

adj_data |>
  distinct(Year, Name, Sex) |>
  count(Year, Sex) |>
  ggplot(aes(x = Year, y = n, colour = Sex)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  labs(x = "Year", y = "Number of Athletes", colour = "Sex") +
  theme_minimal()
```
:::

::: column
```{r}
#| label: fig-topcountries
#| fig-cap: "Top 10 Countries by Total Medal Count in the Summer Olympics (1896–2016)"
#| echo: false
#| warning: false

adj_data |>
  filter(!is.na(Medal)) |>
  count(Team, name = "total_medals") |>
  arrange(desc(total_medals)) |>
  head(10) |>
  mutate(Team = fct_reorder(Team, total_medals)) |>
  ggplot(aes(x = total_medals, y = Team)) +
  geom_col(fill = "steelblue") +
  labs(x = "Total Medals", y = "Country") +
  theme_minimal()
```
:::
:::::

## Visualisation

```{r}
#| label: tbl-medals
#| tbl-cap: "Top 5 Countries by Total Medals (Summer Olympics)"
#| echo: false
#| warning: false
adj_data |>
  filter(!is.na(Medal)) |>
  count(Team, name = "Total Medals") |>
  arrange(desc(`Total Medals`)) |>
  head(5) |>
  rename(Country = Team) |>
  kable(booktabs = TRUE)
```

## Overall participation patterns over time

```{r}
#| include: false
participation <- adj_data_c %>%
  distinct(ID, Year, Season) %>%
  group_by(Year) %>%
  summarise(
    Athletes = n(),
    .groups = "drop"
  )

participation_model <- lm(Athletes ~ Year,
                          data = participation)
```

```{r}
#| echo: false
#| label: tbl-participation
participation_summary <- tibble(
  Term = c("Intercept", "Year"),
  Coefficient = coef(participation_model),
  `p-value` = summary(participation_model)$coefficients[, "Pr(>|t|)"]
)

participation_summary$`R²` <- c(
  summary(participation_model)$r.squared,
  ""
)

kable(participation_summary, digits = 4)
```

```{r}
#| label: fig-participation-residuals
#| fig-cap: "Residual plots for participation model"
#| echo: false
#| fig-width: 5
#| fig-height: 2.5
par(mfrow = c(1,2), mar = c(4,4,1,1))

plot(participation_model, which = 1, main = "")
plot(participation_model, which = 2, main = "")

par(mfrow = c(1,1))
```

## Female Medal Success over time

```{r}
#| include: false
medals <- adj_data_c %>%
  filter(!is.na(Medal))
gender_medals <- medals %>%
  group_by(Year, Sex) %>%
  summarise(
    Medals = n(),
    .groups = "drop"
  ) %>%
  tidyr::pivot_wider(
    names_from = Sex,
    values_from = Medals,
    values_fill = 0
  ) %>%
  rename(
    FemaleMedals = F,
    MaleMedals = M
  )

female_medal_model <- glm(
  cbind(FemaleMedals, MaleMedals) ~ Year,
  family = binomial,
  data = gender_medals
)
```

```{r}
#| echo: false
#| label: tbl-female-medal-success

female_medal_summary <- tibble(
  Term = c("Intercept", "Year"),
  Coefficient = coef(female_medal_model),
  `p-value` = summary(female_medal_model)$coefficients[, "Pr(>|z|)"]
)

female_medal_summary$`Pseudo R²` <- c(
  1 - female_medal_model$deviance /
    female_medal_model$null.deviance,
  ""
)

kable(female_medal_summary, digits = 4)
```

```{r}
#| label: fig-female-medal-residuals
#| fig-cap: "Residual plots for female medal success model"
#| echo: false
#| fig-width: 5
#| fig-height: 2.5
par(mfrow = c(1,2), mar = c(4,4,1,1))

plot(female_medal_model, which = 1, main = "")
plot(female_medal_model, which = 2, main = "")

par(mfrow = c(1,1))
```

## Medal Concentration Among Nations

```{r}
#| include: false
country_medals <- medals %>%
  group_by(Year, NOC) %>%
  summarise(
    Medals = n(),
    .groups = "drop"
  )

hhi_data <- country_medals %>%
  group_by(Year) %>%
  mutate(
    Share = Medals / sum(Medals)
  ) %>%
  summarise(
    HHI = sum(Share^2),
    .groups = "drop"
  )

hhi_model <- lm(HHI ~ Year,
                data = hhi_data)
```

```{r}
#| echo: false
#| label: tbl-hhi

hhi_summary <- tibble(
  Term = c("Intercept", "Year"),
  Coefficient = coef(hhi_model),
  `p-value` = summary(hhi_model)$coefficients[, "Pr(>|t|)"]
)

hhi_summary$`R²` <- c(
  summary(hhi_model)$r.squared,
  ""
)

kable(hhi_summary, digits = 4)
```

```{r}
#| label: fig-hhi-residuals
#| fig-cap: "Residual plots for HHI model"
#| echo: false
#| fig-width: 5
#| fig-height: 2.5
par(mfrow = c(1,2), mar = c(4,4,1,1))

plot(hhi_model, which = 1, main = "")
plot(hhi_model, which = 2, main = "")

par(mfrow = c(1,1))
```

## Conclusion and Recommendation

- Female representation among Olympic athletes has increased (visualisation), and the proportion of medals won by female athletes has also increased significantly over time

-   The distribution of Olympic medals across nations has become less concentrated over time

**Limitation:** Not account for other factors and the linear and logistic regression models not fully capture non-linear changes

**Future research:** : Explore more complex models and conduct qualitative analyses to understand the social and political factors driving changes in Olympic participation and success
